---
title: "Thesis Draft 1"
subtitle: "Joint prior distributions for hierarchical phylogenetic models "
author:
  - Yordan Saputra
# date: 03/31/2026
# date-format: long
filters:
  - diagram
format: 
  html:
    toc: true
    toc-expand: true
    # include-in-header:
      # text: |
        # \usepackage{amsmath}
    grid:
      sidebar-width: 200px
      body-width: 1030px
editor: visual
# bibliography: references.bib
reference-section-title: References
---

# Problem Description

Phylogenetic trees are used to represent the evolutionary history between a set of species. The implied dependencies between species can be expressed statistically via phylogenetic models. However, these models are often hard to estimate and may require strong prior knowledge, which we express in Bayesian statistics via prior distributions. The goal of this thesis is to implement and evaluate joint prior distributions for the variance parameters of phylogenetic models. We have concrete ideas for such priors, which we will discuss with you upon starting your thesis.

# Project Structure

-   Literature search on priors for phylogenetic models
-   Implement the new and existing priors in Stan
-   Run simulation studies
-   Analyse and discuss results

# Libraries & Helper Function

```{r}
#| warning: false
#| message: false

library(ape)
library(brms)
library(cmdstanr)
library(corrplot)
library(ggplot2)
library(ggplotify)
library(gridExtra)
library(knitr)
library(posterior)
library(rstan)

# check_cmdstan_toolchain()
# install_cmdstan(cores = 2)

SEED = 67
set.seed(67)

load_or_fit_brms <- function (data, model_name, formula, family, data2, prior, replace = FALSE) {
  modelpath <- file.path(getwd(), 'PhyloModel', paste0(model_name, '.rds'))
  
  if (file.exists(modelpath) && !replace) {
    model <- readRDS(modelpath)
  } else {
    file.remove(modelpath)
    model <- brm(
      formula, data = data, family = family, 
      data2 = data2, prior = prior, 
      refresh = 0, silent = 2, 
      control = list(adapt_delta = 0.95),
      seed = SEED
    )
    saveRDS(model, modelpath)
  }
  
  return(model)
}

load_or_fit_stan <- function (data, model_name, stan_model, replace = FALSE) {
  modelpath <- file.path(getwd(), 'PhyloModel', paste0(model_name, '.rds'))
  stanpath <- file.path(getwd(), 'PhyloModel', paste0(model_name, '.stan'))
  exepath <- file.path(getwd(), 'PhyloModel', paste0(model_name, '.exe'))
  
  if (file.exists(modelpath) && !replace) {
    model <- readRDS(modelpath)
  } else {
    file.remove(c(modelpath, stanpath, exepath))
    writeLines(stan_model@model_code[1], stanpath)
    stan_model <- cmdstan_model(stanpath)
    
    model <- stan_model$sample(
      data = list(
        N = nrow(data), y = data$phen,
        x1 = data$spec_mean_cf, x2 = data$within_spec_cf,
        J = length(unique(data$species)),
        group = as.numeric(factor(data$species)),
        A_phylo = A
      ),
      chains = 4, parallel_chains = 4,
      iter_warmup = 1000, iter_sampling = 1000,
      adapt_delta = 0.95, max_treedepth = 12,
      seed = SEED,
    )
    model$save_object(file = modelpath)
  }
  
  return(model)
}
```

# Loading Dataset

For now, let's use the subset of the data for faster fitting.

```{r}
#| warning: false

phylo <- ape::read.nexus("https://paul-buerkner.github.io/data/phylo.nex")
data <- read.table("https://paul-buerkner.github.io/data/data_repeat.txt", header = TRUE)
subset_species <- sample(unique(data$species), 15)
data <- data[data$species %in% subset_species, ]
data$spec_mean_cf <- with(data, sapply(split(cofactor, phylo), mean)[phylo])
data$within_spec_cf <- data$cofactor - data$spec_mean_cf

species_to_keep <- data$phylo
phylo <- ape::keep.tip(phylo, species_to_keep)
A <- ape::vcv.phylo(phylo)
```

```{r}
#| echo: false
#| warning: false

rmarkdown::paged_table(data, options = NULL)
```

```{r, fig.height=6, fig.width=12}
#| echo: false
#| warning: false

p1 <- as.grob(~ corrplot(A, method = "color", is.corr = FALSE))
p2 <- as.grob(~ plot(phylo))

grid.arrange(p1, p2, ncol = 2)
```

# Learn Stan

{{< include model_diagram.qmd >}}

{{< include model_stan.qmd >}}

{{< include model_brms.qmd >}}

# MultiPhylo vs. BYM2Params

```{r}
#| echo: false
#| warning: false

# var
mp_essb_phy = subset(fstan_multiphylo_summary, variable == "sigma_phy")$ess_bulk / fstan_multiphylo_time$total
mp_esst_phy = subset(fstan_multiphylo_summary, variable == "sigma_phy")$ess_tail / fstan_multiphylo_time$total
mp_essb_sp = subset(fstan_multiphylo_summary, variable == "sigma_sp")$ess_bulk / fstan_multiphylo_time$total
mp_esst_sp = subset(fstan_multiphylo_summary, variable == "sigma_sp")$ess_tail / fstan_multiphylo_time$total
bym2_essb_phy = subset(fstan_bym2_summary, variable == "sigma_phy")$ess_bulk / fstan_bym2_time$total
bym2_esst_phy = subset(fstan_bym2_summary, variable == "sigma_phy")$ess_tail / fstan_bym2_time$total
bym2_essb_sp = subset(fstan_bym2_summary, variable == "sigma_sp")$ess_bulk / fstan_bym2_time$total
bym2_esst_sp = subset(fstan_bym2_summary, variable == "sigma_sp")$ess_tail / fstan_bym2_time$total

ess_per_sec <- data.frame(
  variable = c("sigma_phy", "sigma_sp"),
  multiphylo_bulk = c(mp_essb_phy, mp_essb_sp),
  multiphylo_tail = c(mp_esst_phy, mp_esst_sp),
  bym2params_bulk = c(bym2_essb_phy, bym2_essb_sp),
  bym2params_tail = c(bym2_esst_phy, bym2_esst_sp)
)
```

```{r}
#| echo: false
#| warning: false

kable(ess_per_sec)
```

# Scratch Paper



